\chapter{Methodology and testbed architecture}
\label{ch:methodology}

\begin{quote}\itshape
This chapter in brief: the project is six chained phases, each ending in a pass/fail gate
(\cref{fig:pipeline}). Four safeguards keep the AI-versus-traditional comparison honest;
one section explains every machine-learning tool by name before it is used; and the software
inventory maps each program to its role. Read \cref{sec:ai-toolbox} even if you skip the rest.
\end{quote}

\keyidea{Credibility comes from pre-registration: hypotheses, thresholds and tests are frozen before the test data is seen, and a failed gate triggers redesign or an honest refutation, never a quiet threshold move. A project that cannot fail its own gates is advertising, not research.}


\section{Overall architecture}

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[font=\footnotesize, node distance=3.5mm,
      ph/.style={draw, rounded corners=1pt, minimum height=9mm, minimum width=17mm,
                 align=center, fill=blue!8},
      gate/.style={font=\scriptsize, text=black!60}]
    \node[ph] (f1) {F1\\engine twin};
    \node[ph, right=of f1] (f2) {F2\\synthetic fleet};
    \node[ph, right=of f2] (f3) {F3\\traditional EHM};
    \node[ph, right=of f3] (f4) {F4\\AI EHM};
    \node[ph, right=of f4] (f5) {F5\\evaluation};
    \node[ph, right=of f5] (f6) {F6\\cases + report};
    \foreach \a/\b/\g in {f1/f2/H1, f2/f3/H2, f3/f4/H3, f4/f5/H4, f5/f6/H5}{
      \draw[-{Stealth}] (\a) -- (\b) node[midway, above, gate] {\g};}
    \node[gate, right=2mm of f6] {H6};
    \node[gate, below=1.5mm of f1] {decks + ICM};
    \node[gate, below=1.5mm of f2] {SynCFM56};
    \node[gate, below=1.5mm of f3] {baseline scores};
    \node[gate, below=1.5mm of f4] {AI scores};
    \node[gate, below=1.5mm of f5] {verdicts H1--H5};
  \end{tikzpicture}
  \caption{The six project phases, their gates (H1--H6) and their key artifacts. Everything to
  the left of an arrow must pass its gate before the next phase starts.}
  \label{fig:pipeline}
\end{figure}

The testbed is organized in six chained phases, each closed by a milestone (\emph{gate}) with a
measurable acceptance criterion:

\begin{enumerate}
  \item[\textbf{F1.}] \textbf{Thermodynamic twin} of the CFM56-7B26 in pyCycle: design point,
    off-design behavior, baseline decks over the flight envelope (pre-computed tables of what a
    healthy engine reads at every operating condition) and the influence coefficient matrix.
    Gate H1: error $\leq 3$\,\% against certified data at the anchor points.
  \item[\textbf{F2.}] \textbf{Synthetic fleet generator} (SynCFM56): per-mechanism health
    trajectories, a hierarchical fleet of $\sim$100 engines run to failure, a sensor model
    (noise, bias, quantization, missing data) and ACARS-style snapshots with ground truth.
    Gate H2: dataset audited for realism and difficulty, no leakage between data splits, and a
    written datasheet.
  \item[\textbf{F3.}] \textbf{Traditional EHM} reference: baselines and smoothed deviations,
    trend monitoring with persistence rules, WLS GPA (\cref{eq:wls}), Kalman GPA
    (\cref{eq:kalman}), expert-rule fault isolation, and RUL by robust extrapolation of the EGT
    margin. Gate H3: full metrics on the test set.
  \item[\textbf{F4.}] \textbf{AI-based EHM}: anomaly detection (autoencoders, Isolation
    Forest), multi-class fault diagnosis (gradient boosting, neural networks), RUL prognosis
    (LSTM/TCN/Transformer sequence models), the physics-informed hybrid, conformal uncertainty
    and physical explainability (PCS). Gate H4.
  \item[\textbf{F5.}] \textbf{Pre-registered comparative evaluation}: common metrics,
    statistical tests, ablations, and sim-to-real validation on N-CMAPSS. Gate H5.
  \item[\textbf{F6.}] Case studies, interactive dashboard, and final writing. Gate H6.
\end{enumerate}

\section{Machine learning without prior background}
\label{sec:ml-primer}

Nothing in this document assumes machine-learning training. Six ideas carry everything:

\begin{description}
  \item[Learning from data.] A machine-learning model is a mathematical function with many
    adjustable numbers inside (its \emph{parameters}). ``Training'' means showing it examples
    (here: deviation histories with known outcomes, because the fleet is synthetic) and
    letting an algorithm nudge those numbers until the function's outputs match the known
    answers. Nobody programs the rules; the rules emerge from the examples.
  \item[Train / validation / test.] Three disjoint groups of engines with three jobs.
    \emph{Training} engines are the textbook the model studies. \emph{Validation} engines
    are practice exams: used to choose between model variants and settings, never to adjust
    parameters directly. \emph{Test} engines are the final exam, sat exactly once --- any
    peeking turns the grade into fiction. The same discipline is applied to the traditional
    pipeline's thresholds.
  \item[Overfitting.] A model with enough capacity can memorize its textbook --- perfect on
    training engines, useless on new ones. That is why every number that matters in this
    document comes from engines the models never saw, and why splitting by \emph{engine}
    (not by flight) is non-negotiable: flights of the same engine are near-duplicates.
  \item[Neural networks, in one paragraph.] A neural network is a chain of simple layers:
    each layer multiplies its inputs by adjustable weights, sums them, and passes the result
    through a mild nonlinearity. Stacked layers can approximate very general input--output
    relationships. Training adjusts the weights by \emph{gradient descent}: compute how
    wrong the output is, compute in which direction each weight should move to reduce that
    error, take a small step, repeat over the dataset for a number of \emph{epochs}
    (passes). A \emph{recurrent} network (like the GRU used here) additionally carries a
    small memory from one time step to the next, so it can read a sequence --- an engine's
    deviation history --- the way a reader accumulates context along a sentence.
  \item[Hyperparameters and tuning.] Settings the training process itself does not adjust
    --- network size, learning-rate, window lengths, alert thresholds --- are
    \emph{hyperparameters}. Choosing them by trial and error on validation data is
    \emph{tuning}; automating that search is what Optuna does. Both EHM families received
    exactly 50 tuning trials (\cref{sec:safeguards}): the comparison is between tuned
    practitioners, not between a tuned one and a default one.
  \item[Seeds and variance.] Training involves randomness (initial weights, sampling
    order). A \emph{seed} fixes that randomness so runs are repeatable; training with
    several seeds and reporting the spread separates a real effect from a lucky draw ---
    which is exactly how the hybrid experiment's negative result was established.
\end{description}

\section{The machine-learning toolbox, briefly}
\label{sec:ai-toolbox}

With those six ideas in hand, the specific techniques, one paragraph each:

\begin{description}
  \item[Autoencoder (anomaly detection).] A neural network trained to compress and then
    reconstruct \emph{healthy} data. Fed a degraded snapshot, it reconstructs it poorly; the
    reconstruction error is the anomaly score. No fault labels needed.
  \item[Isolation Forest (anomaly detection).] An ensemble of random trees; anomalous points
    are isolated in fewer random splits than normal ones. A classical non-neural baseline.
  \item[Gradient boosting / XGBoost (diagnosis).] An ensemble of small decision trees, each
    correcting the errors of the previous ones. State of the art on tabular data; predicts the
    fault class of a snapshot window.
  \item[LSTM, TCN, Transformer (RUL prognosis).] Three sequence-model families ---recurrent,
    convolutional and attention-based--- that consume a sliding window of past snapshots and
    regress the remaining useful life. The v0 suite implements a GRU (recurrent); the
    cross-family comparison belongs to the F5 tuned round.
  \item[Conformal prediction (uncertainty).] A distribution-free wrapper: from the model's
    errors on a calibration set it builds prediction intervals with a mathematically
    guaranteed coverage rate (e.g.\ 90\,\%), regardless of the underlying model.
  \item[SHAP (explainability).] Assigns each input feature its contribution to one specific
    prediction (a game-theoretic attribution). The PCS (contribution C5) projects these
    attributions through the ICM into health-parameter space to test their physical coherence.
  \item[Optuna (tuning).] An automated hyperparameter search library; a ``trial'' is one
    training run with a candidate configuration. Both EHM families get the same trial budget.
\end{description}

Supporting infrastructure named later: \emph{DVC} versions datasets the way git versions code;
\emph{MLflow} records every training run (configuration, metrics, artifacts) so experiments
are traceable and repeatable.

\section{Designing a fair comparison, as method}
\label{sec:fair-design}

The comparison itself is an instrument, and it was designed before any contestant ran.
Three principles, stated as axioms because every later choice traces to one of them:

\begin{description}
  \item[Symmetry.] Both families see \emph{identical} inputs (same snapshots, same
    deviations against the same published baseline), are scored by \emph{identical}
    metrics on \emph{identical} splits, and receive \emph{identical} optimization
    budgets (50 tuning trials each, \cref{sec:stats-math}). Any asymmetry --- richer
    features for one side, hand-tuning for the other --- silently converts a method
    comparison into an effort comparison. The literature's chronic disease
    (\cref{ch:introduction}) is exactly that conversion.
  \item[Evidence hierarchy.] Numbers exist in two castes, never mixed: \emph{exploratory}
    (produced while designing, debugging, iterating --- chapters 6--7) and
    \emph{confirmatory} (produced once, after a frozen pre-registration, on data touched
    exactly once --- chapters 8 and 10). Only the second caste supports verdicts. The
    discipline is borrowed from clinical trials, where the cost of self-deception is
    measured in lives; here it is merely measured in wrong fleet decisions.
  \item[Gates over milestones.] Progress is declared by \emph{passing pre-stated
    acceptance criteria}, not by finishing work. Gates are allowed to fail (the
    difficulty audit failed twice, hypothesis H7.3 died on its frozen thresholds), and a
    failed gate triggers redesign or an honest refutation --- never a quiet threshold
    adjustment. A project that cannot fail its own gates is advertising, not research.
\end{description}

\section{The statistical machinery, in full}
\label{sec:stats-math}

Every verdict in this document rests on the following tests. Each is stated with its
mathematics and the reason it was chosen; nothing later in the document uses a statistic
not defined here (norm N6).

\paragraph{Wilcoxon signed-rank test (paired continuous outcomes).} For paired errors
$(a_i, b_i)$ on the same $n$ engines, compute differences $\delta_i = a_i - b_i$, discard
zeros, rank $|\delta_i|$ ascending, and sum the ranks of the positive differences:
\begin{equation}
  W \;=\; \sum_{i:\,\delta_i > 0} \operatorname{rank}\bigl(|\delta_i|\bigr).
  \label{eq:wilcoxon}
\end{equation}
Under the null (differences symmetric about zero) every rank is equally likely to carry
either sign, giving an exact, distribution-free reference: no normality assumption, which
matters because RUL errors are heavy-tailed. Pairing by engine removes engine-to-engine
difficulty variation from the comparison --- the same reason a crossover trial beats two
separate cohorts. \emph{Worked example}: $n=6$ engines with differences
$\{+3, +8, -2, +5, +9, +1\}$: ranks of magnitudes $\{3,7\ldots\}$ --- the two negative
values carry little rank mass, $W$ lands near its maximum, and the one-sided exact
$p \approx 0.047$.

\paragraph{McNemar's exact test (paired binary outcomes).} For paired correct/incorrect
outcomes (e.g.\ both families isolating the same episodes), only the discordant pairs are
informative: $b$ = cases only method A got right, $c$ = only method B. Under the null of
equal skill each discordant pair is a fair coin, so
\begin{equation}
  p \;=\; \Pr\bigl[X \ge b\bigr], \qquad X \sim \mathrm{Binomial}(b+c,\ \tfrac12),
  \label{eq:mcnemar}
\end{equation}
one-sided when the hypothesis is directional. The concordant cases cancel --- episodes both
get right (or both miss) say nothing about the \emph{difference}. This is why chapter 10
could bound its verdict without replaying the test pass: with 15/23 vs 4/23 correct, the
accuracy gap forces $b - c = 11$, and $p$ is maximized at $b=15, c=4$, still $0.0096$.

\paragraph{Holm correction (multiple comparisons).} Testing $m$ hypotheses at
$\alpha = 0.05$ each inflates the chance of at least one false confirmation. Holm's
step-down procedure sorts the p-values $p_{(1)} \le \cdots \le p_{(m)}$ and compares
$p_{(k)}$ against $\alpha / (m - k + 1)$, stopping at the first failure:
\begin{equation}
  p^{\text{Holm}}_{(k)} \;=\; \min\!\Bigl(1,\ \max_{j \le k}\ (m - j + 1)\, p_{(j)}\Bigr).
  \label{eq:holm}
\end{equation}
It controls the family-wise error rate at $\alpha$ with no independence assumption, and is
uniformly more powerful than Bonferroni (which would multiply every $p$ by $m$).

\paragraph{BCa bootstrap confidence intervals.} Resample engines (the exchangeable unit)
with replacement $B = 10^4$ times, recompute the statistic $\hat\theta^{*}_b$, and read the
interval from the bootstrap distribution --- but with two corrections: the \emph{bias}
term $z_0 = \Phi^{-1}\bigl(\#\{\hat\theta^{*}_b < \hat\theta\}/B\bigr)$ (is the bootstrap
distribution off-center?) and the \emph{acceleration} $a$ estimated by jackknife (does the
statistic's variance depend on its value?). The adjusted percentiles
\begin{equation}
  \alpha_{1,2} \;=\; \Phi\!\Bigl(z_0 + \tfrac{z_0 + z_{\alpha_{1,2}}}{1 - a\,(z_0 + z_{\alpha_{1,2}})}\Bigr)
  \label{eq:bca}
\end{equation}
replace the naive 2.5/97.5 cuts. For small $n$ (20 test engines) and skewed statistics
(RMSE), the corrections matter --- the naive percentile interval can miss its nominal
coverage badly.

\paragraph{Cliff's delta (effect size).} Significance says an effect exists; $\delta$ says
how big:
\begin{equation}
  \delta \;=\; \frac{\#\{(i,j): a_i > b_j\} - \#\{(i,j): a_i < b_j\}}{n_a\, n_b}
  \;\in\; [-1, 1],
  \label{eq:cliff}
\end{equation}
the probability a random error from method A exceeds a random one from B, minus the
reverse. $\delta = 0.89$ (the H3 verdict) means: pick any traditional engine-error and any
AI engine-error at random, and the traditional one is larger about 94\,\% of the time.
Rank-based, hence robust to the same heavy tails that motivated Wilcoxon.

\paragraph{Split conformal prediction (distribution-free intervals).} Fit any predictor on
the training set; on a held-out calibration set of size $n_c$, collect the absolute
residuals $r_i = |y_i - \hat{y}_i|$ and take
\begin{equation}
  \hat{q} \;=\; \text{the } \lceil (1-\alpha)(n_c + 1)\rceil\text{-th smallest } r_i;
  \qquad
  C(x) \;=\; \bigl[\hat{y}(x) - \hat{q},\ \hat{y}(x) + \hat{q}\bigr].
  \label{eq:conformal}
\end{equation}
Exchangeability alone guarantees $\Pr[y \in C(x)] \ge 1 - \alpha$ --- no distributional
assumption, any model, finite samples. The $(n_c + 1)$ (not $n_c$) accounts for the new
point joining the ranking: the guarantee is exact, not asymptotic. The classification
variant used in chapter 10 (adaptive prediction sets) sorts the predicted class
probabilities, accumulates until the calibration quantile of ``mass needed to reach the
true class'' is covered, and returns that \emph{set} of classes --- ambiguity becomes set
size instead of silent error.

\paragraph{Tree-structured Parzen estimator (the tuner).} Optuna's TPE turns tuning into
inference: after some trials, split configurations into the best fraction ($\ell(\theta)$,
a density fitted over good configs) and the rest ($g(\theta)$), then propose the $\theta$
maximizing $\ell(\theta)/g(\theta)$ --- sampling where good configurations concentrate,
exploring where $g$ is thin. Fifty trials of TPE typically cover a 14--16-dimensional
mixed space far better than grid search's curse of dimensionality allows.

\section{Methodological safeguards}
\label{sec:safeguards}

The credibility of an AI-versus-traditional comparison depends on eliminating the biases that
plague the literature. Four safeguards are adopted:

\begin{description}
  \item[Pre-registration.] Hypotheses (\cref{tab:hypotheses}), metrics, data splits (with file
    hashes), statistical tests and the stopping rule are frozen in the repository under a git
    tag (\code{prereg-v1}, document \code{docs/prereg-v1.md}) \emph{before} the confirmatory
    evaluation. This borrows a standard practice from medicine: you state what would convince
    you before looking at the answer, so you cannot move the goalposts afterwards. The frozen
    document also discloses every exploratory test-set exposure that preceded it.
  \item[Symmetric tuning budget.] Every method has knobs. The traditional pipeline's
    (alert thresholds, regularization $\lambda$, Kalman $\mathbf{Q}$, smoothing windows) and
    each AI family's hyperparameters are tuned with the \emph{same number} of automated search
    trials (Optuna), using only the training and validation engines.
  \item[Split by engine.] The fleet is divided 70/10/20 engines into training, validation and
    test sets, and no engine ever appears in two sets. Splitting by flight instead of by engine
    ---a common mistake--- would let a model memorize an engine's individual quirks and score
    unrealistically well. An automated leakage check runs in continuous integration.
  \item[A strong baseline.] The traditional pipeline is implemented with the same care as the
    AI, every expert rule documented with its physical justification; smearing is
    \emph{measured} against ground truth, not assumed.
\end{description}

\section{Reproducibility and infrastructure}
\label{sec:reproducibility}

The entire project is a public
repository\footnote{\url{https://github.com/cedarcreeks/EHMbrAIn}} with a reproducible
environment: Python~3.11 managed with \code{uv} and a fully pinned dependency lockfile;
pyCycle~4.4 on OpenMDAO~\cite{Gray2019}; declarative YAML configuration; and automated tests
(pytest) executed by continuous integration ---a service that re-runs the whole test suite on
every change, so a regression cannot land silently. Data and experiments of later phases will
be versioned with DVC and MLflow respectively.

Three concrete decisions illustrate the intended standard of rigor:

\begin{itemize}
  \item \textbf{A versioned calibration contract.} The model's numerical targets
    (\cref{tab:tcds}) live in a single file (\code{conf/cfm56_7b_targets.yaml}) carrying
    value, tolerance, source and status (\code{VERIFIED}/\code{TO_VERIFY}) per entry. Both the
    model runners and the tests read that file: changing a target is a visible version-control
    event, never a silent edit.
  \item \textbf{Dependency canary tests.} During environment setup, an incompatibility between
    pyCycle~4.4 and numpy~$\geq 2$ was discovered (a failure inside the tabular thermodynamics
    package). The fix pins \code{numpy<2}, and a smoke test that runs a complete engine cycle
    executes on every CI run, so any future dependency upgrade that breaks the solver fails
    loudly and immediately.
  \item \textbf{Auto-generated evidence.} Every result figure and table in this document is
    produced by a script (\code{scripts/make_report_assets.py}) that solves the model and
    writes the PDFs, the \LaTeX{} table files (\code{paper/report/tables/*.tex}) and a
    provenance file (\code{assets.json}). No result number is ever copied by hand.
  \item \textbf{Compute-efficiency norms.} The repository carries binding engineering norms
    (\code{docs/engineering-norms.md}): long decomposable computations run in parallel across
    CPU cores (thermodynamic sweeps and ICM columns use one worker process per core --- these
    are scalar Newton solves that gain nothing from a GPU), while the tensor workloads of phase
    F4 (network training, SHAP) run on the GPU (Apple-silicon MPS backend on the development
    machine).
\end{itemize}

\section{Software architecture and computational pipeline}
\label{sec:software}

This section is the complete inventory of the project's code: what each program does, how it
does it, and what flows between them. It is updated whenever code lands.

\subsection{Library modules (\texttt{src/ehmbrain/})}

\begin{description}
  \item[\code{perf/cycle.py} --- the engine model.] Defines four model classes and their
    builders.
    \begin{itemize}
      \item \code{CFM56(pyc.Cycle)}: one thermodynamic point. Builds the component graph
        (inlet, fan, splitter, booster, HPC with three bleed ports, post-compressor bleed,
        combustor, cooled HPT/LPT, ducts, two convergent nozzles, two shafts, performance
        block) and its balance system. In \emph{design} mode the four balances solve mass flow
        for thrust, fuel--air ratio for $T_{t4}$, and both turbine expansion ratios for zero
        net shaft power. In \emph{off-design} mode they solve mass flow and bypass ratio to
        hold the design nozzle throat areas, both spool speeds for shaft power balance, and
        fuel--air ratio according to the throttle mode: \code{T4} (rating temperature),
        \code{percent_thrust} (a fraction of a reference thrust) or \code{N1} (hold fan
        speed, the CFM56 rating parameter, implemented through a passthrough component because
        the speed is itself another balance's output). Each point carries its own Newton
        solver (tolerance $10^{-8}$, Armijo--Goldstein line search, subsystem solves enabled).
      \item \code{MPCFM56(pyc.MPCycle)}: DESIGN + the four SLS anchors A2--A5 as
        thrust-matched \code{percent_thrust} points. Used for calibration
        (\cref{sec:calibration}).
      \item \code{MPCFM56Study}: DESIGN + one off-design point whose five turbomachine map
        scalars (efficiency and flow) pass through multiplier components
        (\code{health_<comp>.eta_fac/flow_fac}); setting a factor to $1+\Delta$ imposes a
        health deviation. Used for the ICM and, in F2, for degraded-fleet snapshots.
      \item \code{MPCFM56Deck}: DESIGN + \code{OD_max} (maximum thrust at the rating $T_4$)
        + \code{OD_pt} (a commanded fraction of that maximum). Used for the baseline decks.
      \item Builders (\code{build_problem}, \code{build_study_problem},
        \code{build_deck_problem}) return solver-ready problems with all design inputs and
        balance initial guesses set; \code{solve_anchors()} encapsulates the idle power
        continuation ($0.30\rightarrow0.07$); \code{snapshot()} reads the full sensor set at a
        point; \code{set_health()} imposes a health-deviation dictionary.
    \end{itemize}
  \item[\code{perf/icm.py} --- GPA sensitivity analysis.] \code{compute_icm()} builds a Study
    problem per health parameter (in parallel worker processes, norm N1), perturbs it
    $\pm0.5\,\%$ at constant N1 and forms the central-difference column in \%-per-\%;
    \code{svd_report()} returns rank, singular values and condition number;
    \code{signature_angles()} the pairwise angles between fault signatures; and
    \code{confusable_pairs()} the sub-$15^{\circ}$ subset that instantiates hypothesis H2.
  \item[\code{common/corrections.py} --- corrected parameters.] ISA atmosphere
    (\code{isa_static}), ram-corrected $\theta/\delta$ at the fan face
    (\code{theta_delta}), the exponent regression \code{fit_correction_exponents()}
    ($\log Y$ against a cubic polynomial in log corrected speed plus $a\log\theta + b\log
    \delta$, solved by least squares) and \code{correct()} to apply the fitted law.
\end{description}

\subsection{Executable scripts (\texttt{scripts/})}

Each script is a pipeline stage writing versioned artifacts under \code{data/processed/}:

\begin{center}\small
\begin{tabular}{p{0.27\textwidth}p{0.66\textwidth}}
  \toprule
  Script & Function and outputs \\
  \midrule
  \code{run_design_point.py} & Solves the design point, checks the five WP1.1 acceptance
    windows against the contract; nonzero exit on failure. \\
  \code{run_anchors.py} & Solves DESIGN + A2--A5 with the idle continuation and compares
    fuel-flow/OPR predictions to the EEDB (gated tolerances); prints the verdict table. \\
  \code{make_decks.py} & Envelope sweep for the baseline decks: independent (altitude, Mach)
    blocks in parallel processes, warm continuation inside a block, cold-rebuild recovery
    (norm N2), then a per-channel parallel leave-one-out audit.
    $\rightarrow$ \code{baseline_deck.parquet}, \code{deck_report.json}. \\
  \code{make_corrected_baseline.py} & Fits the correction exponents on the deck and re-runs
    the leave-one-out audit in corrected space. $\rightarrow$
    \code{corrected_baseline.json}. \\
  \code{make_icm.py} & ICM at the six-point operating grid (columns in parallel), SVD and
    confusability analysis. $\rightarrow$ \code{icm_<point>.npz}, \code{icm_report.json}. \\
  \code{cea_crosscheck.py} & Re-solves the design point with CEA thermodynamics and reports
    deltas against tabular. $\rightarrow$ \code{cea_crosscheck.json}. \\
  \code{make_report_assets.py} & Regenerates every figure and table of this document from
    the model and the artifacts above (norm N4). $\rightarrow$
    \code{paper/report/figures/*.pdf}, \code{paper/report/tables/*.tex},
    \code{assets.json}. \\
  \bottomrule
\end{tabular}
\end{center}

Later phases added, under the same contract-artifact-evidence pattern:
\begin{description}
  \item[\code{datagen/}] \code{trajectories.py} (per-mechanism health profiles, exact wash
    sawtooth, multi-episode acute faults), \code{fleet.py} (engine sampling, EGT-margin end
    of life, leakage-free splits), \code{sensors.py} (noise/quantization/drift/dropout),
    \code{snapshots.py} (ACARS-style two-report generator via the F1 linearization).
    Driver: \code{make_fleet.py}; audits: \code{audit_nonlinearity.py},
    \code{audit_dataset.py}.
  \item[\code{trad/}] \code{pipeline.py}: baseline deviations, Holt smoothing with
    innovations, CUSUM and dual-EWMA gap detectors, WLS and wash-aware Kalman GPA,
    nearest-signature isolation, Theil--Sen RUL. Driver: \code{run_trad.py}.
  \item[\code{ai/}] \code{data.py} (shared feature builder --- same deviations as
    \code{trad/}), \code{models.py} (autoencoder, GRU RUL net, MPS-aware training loops).
    Drivers: \code{run_ai.py} (detection/diagnosis/RUL + conformal),
    \code{run_hybrid.py} (stacking ablation), \code{run_pcs.py} (Physics-Consistency
    Score), \code{tune_detection.py} (validation sweep).
\end{description}

The data flow is linear: \code{conf/*.yaml} (the contracts) feed the model and generator;
runners and \code{make_*} scripts write \code{data/processed/*} artifacts;
\code{make_report_assets.py} turns artifacts into report evidence; the tests gate every
stage.

\subsection{Test suite (\texttt{tests/})}

Thirty-six tests, all run in CI on every change: environment smoke (the vendored pyCycle
example end-to-end); design-point convergence and acceptance windows; anchor convergence,
fuel-flow and OPR tolerances, and N1/N2 redlines; the Study self-consistency property; ICM
physical signs and rank-3 cockpit underdetermination; the corrections module (ISA,
$\theta/\delta$, exponent-fit recovery on synthetic truth); trajectory physics (wash
sawtooth ratios and regrowth, FOD steps, reproducibility, split integrity); the sensor and
snapshot models (noise statistics, quantization, drift ramps, dropout fraction, end-to-end
consistency); traditional-pipeline primitives (Holt trend and innovations, CUSUM
step-not-trend, gap-detector ramp capture, WLS recovery, Kalman observation-space
convergence, Theil--Sen); and AI primitives (feature fill, window alignment, device
selection, autoencoder separation, GRU countdown learning).

\section{Replication guide}
\label{sec:replication}

This document and the repository are designed to be used together: every result chapter
names the script that produced its evidence, and this section gives the complete replication
path. \textbf{Reference machine for every wall-clock figure in this document}: an Apple M5
desktop chip (10 cores, 16\,GB unified memory), with tensor training on its GPU through the
torch MPS backend --- consumer hardware, deliberately. Total cost: about \textbf{11 minutes}
end to end (\cref{tab:compute}; per-stage breakdown measured by
\code{scripts/benchmark_pipeline.py}, norm N5).

\paragraph{Setup (once).}
\begin{enumerate}
  \item Clone \code{github.com/cedarcreeks/EHMbrAIn} and install
    \code{uv} (the Python package manager).
  \item \code{uv sync} --- reproduces the exact pinned environment (Python 3.11,
    pyCycle 4.4, torch with MPS).
  \item \code{uv run pytest} --- 36 tests must pass before anything else; they gate every
    claim in this document.
\end{enumerate}

\paragraph{Pipeline (in order).} \code{make all} runs everything below and rebuilds this
document (gate H6); the stage-by-stage mapping to report evidence is:

\begin{center}\footnotesize
\begin{tabular}{>{\raggedright\arraybackslash}p{0.34\textwidth}>{\raggedright\arraybackslash}p{0.27\textwidth}>{\raggedright\arraybackslash}p{0.30\textwidth}}
  \toprule
  Command (\code{uv run python \ldots}) & Produces & Feeds \\
  \midrule
  \code{scripts/run_design_point.py} & design-point check & \cref{tab:design,tab:stations} \\
  \code{scripts/run_anchors.py} & calibration verdicts & \cref{tab:anchors,fig:anchor-errors} \\
  \code{scripts/make_decks.py} & baseline decks + LOO & \cref{tab:deck-loo} \\
  \code{scripts/make_corrected_baseline.py} & corrected-space exponents & \cref{tab:corrected} \\
  \code{scripts/make_icm.py} & ICM grid + observability & \cref{fig:icm,tab:confusable,tab:icm-grid} \\
  \code{scripts/cea_crosscheck.py} & thermo cross-check & \cref{tab:cea} \\
  \code{scripts/make_fleet.py} & SynCFM56 fleet & \cref{tab:fleet,fig:egtm,fig:trajectory} \\
  \code{scripts/audit_dataset.py} & difficulty + realism audits & \cref{tab:difficulty-audit} \\
  \code{scripts/audit_nonlinearity.py} & linearization audit & \cref{tab:nonlin} \\
  \code{scripts/run_trad.py} & traditional metrics & \cref{tab:trad,fig:kf-smearing,fig:rul-funnel} \\
  \code{scripts/run_ai.py} & AI metrics + conformal & \cref{tab:ai-vs-trad,fig:rul-compare} \\
  \code{scripts/run_hybrid.py} & hybrid ablation & \cref{fig:hybrid} \\
  \code{scripts/run_pcs.py} & PCS attribution study & \cref{sec:pcs} \\
  \code{scripts/tune_f5.py} + \code{scripts/f5_confirm.py} & tuned configs; H1--H5 verdicts & \cref{tab:verdicts} \\
  \code{scripts/sim_to_real.py} & external-benchmark ranking check & \cref{sec:sim-to-real} \\
  \code{scripts/benchmark_pipeline.py} & compute times (norm N5) & \cref{tab:compute} \\
  \code{scripts/make_case_studies.py} & five case-study figures + facts & \cref{ch:cases} \\
  \code{scripts/make_report_assets.py} & \emph{every} figure and table above & this document \\
  \bottomrule
\end{tabular}
\end{center}

Two platform notes a replicator needs (both learned the hard way, both in the decision
register): torch-MPS training runs must execute in the \emph{foreground} on macOS
(backgrounded runs segfault), and XGBoost cannot coexist with torch-MPS in one process
(scikit-learn's histogram gradient boosting is used instead).

\subsection{Compute cost on the reference machine}

\begin{table}[htbp]
  \centering\small
  \caption{Wall-clock time per stage on the reference desktop machine (Apple M5, 10 cores,
  16\,GB, torch MPS backend). Cold-start subprocess timings --- what a replicator actually
  pays. Auto-generated per norm N5.}
  \label{tab:compute}
  \input{tables/table_compute}
\end{table}

The entire evidence base of this document regenerates in 657 seconds --- about eleven
minutes --- on the Apple M5 reference machine, GPU stages included. This is the quantitative
form of transferability rule 4: if it only ran on a cluster, it would not be adoptable.

\section{Comparison metrics}

Both families are scored with identical metrics. For detection: lead time (how many flights
before the ground-truth event the method raises a sustained alert), false alarms per thousand
flights, F1 and precision-recall AUC. For isolation: top-1/top-2 accuracy, macro-F1, a
\emph{smearing index} (the fraction of $\|\hat{\mathbf{x}}\|_1$ assigned to healthy components)
and the PCS. For health estimation: RMSE of $\hat{\mathbf{x}}$ against ground truth. For RUL:
RMSE, MAE, the NASA PHM08 score
\begin{equation}
  S=\sum_i s_i,\qquad
  s_i = \begin{cases} e^{-d_i/13}-1 & d_i<0\\ e^{d_i/10}-1 & d_i\geq 0\end{cases},
  \qquad d_i = \widehat{\mathrm{RUL}}_i - \mathrm{RUL}_i,
\end{equation}
which penalizes late predictions more than early ones (a late removal risks an in-flight
shutdown; an early one merely wastes some life). Two further prognostic metrics need
definitions: \emph{$\alpha$--$\lambda$ accuracy} is the fraction of predictions falling within
$\pm\alpha$ (here 20\,\%) of the true RUL when evaluated at specified life fractions
$\lambda$ (here 50, 70 and 90\,\% of the way to failure) --- it asks ``was the prediction
acceptable at mid-life? near the end?''; the \emph{prognostic horizon} is the earliest moment
from which the prediction stays within the $\pm\alpha$ band for the rest of the engine's life
--- longer horizons mean earlier trustworthy warnings. For uncertainty: empirical versus
nominal interval coverage (does the ``90\,\%'' interval actually contain the truth 90\,\% of
the time?) and mean interval width (narrow honest intervals are the goal; wide intervals are
trivially calibrated but useless).
