\chapter{Conclusions and future work}
\label{ch:conclusions}

\begin{quote}\itshape
This chapter in brief: what was built, what the pre-registered evidence says, what an
adopting organization should do differently on Monday, and what remains honestly open.
\end{quote}

\section{What was built}

From nothing but certified public data, this work produced: a calibrated thermodynamic twin
of the CFM56-7B26 whose fuel-flow predictions sit within 3\,\% of certification test-bed
measurements (C1's foundation); \textbf{SynCFM56}, a 100-engine run-to-failure fleet with
complete per-flight ground truth in airline snapshot format, gated by three audits that
failed and were fixed twice (C1); a traditional EHM pipeline built to industrial shape and a
matched AI suite consuming identical inputs (C2's fairness); conformal uncertainty for
prognosis (C4); the Physics-Consistency Score machinery (C5); the sensor-set observability
analysis (C6's core); an external-benchmark ranking check (C7); and a pre-registered,
Holm-corrected adjudication of five hypotheses under a symmetric 50-trial tuning budget (C2)
--- all reproducible on one desktop machine in minutes and documented to be readable without
prior specialization.

\section{What the evidence says}

Three confirmations, two refutations (\cref{tab:verdicts}), and the split is the message:

\begin{itemize}
  \item \textbf{Learning wins where patterns accumulate}: prognosis (3--6$\times$ lower RUL
    error, calibrated 6$\times$-narrower intervals, ranking replicated on C-MAPSS) and ---
    after honest tuning --- detection timing (12$\times$ earlier at equal false alarms).
  \item \textbf{Physics wins where information is scarce}: the ICM rule matched or beat the
    learner on confusable isolation, and bolting physics-derived features onto a learner
    (H4) made it worse. Physics' real contribution was \emph{diagnosis of the limits}: the
    influence-matrix geometry predicted, before any learning ran, exactly where every method
    would fail.
  \item \textbf{Sensors bound everyone}: the confusable wall (H2) is informational, not
    algorithmic. The extended gas-path set quintuples the worst signature angle; no model
    recovers what probes never measured.
\end{itemize}

\section{Guidance for adoption}

The five-point checklist of \cref{sec:res-industry} stands as the operational summary:
prognosis first, tuned change-detection second, sensors before models for isolation,
distrust of untuned comparisons, and pre-registration as a procurement demand. To it the
case studies add a sixth: \textbf{protect the sensor estate} --- the drifting thermocouple
of Case C corrupts every consumer downstream, learned or classical.

\section{Honest limitations}

The world here is synthetic and linearized (audited to the noise floor, but still a model
of a model); the mission mix is two snapshot conditions with scatter; the EGT is a
station-4.5 proxy; chronic labels correlate with age by construction; the AI family is one
compact architecture per task, not the state of the art; and H4's verdict covers one of
three planned hybrid mechanisms. None of these caveats is hidden in fine print --- each is
in the decision register or the relevant chapter.

\section{Future work}

(1) The two untested hybrid mechanisms --- twin-residual features and physics-constrained
losses --- under the same pre-registered discipline; (2) the PCS evaluated on a competent
classifier, now that fleet v1.1 removes the sample-starvation excuse; (3) N-CMAPSS DS02 as
the heavyweight sim-to-real extension; (4) mission-mix and transient realism in the
generator; (5) the extended-sensor fleet, turning the C6 observability analysis into a
measured acquisition trade study; (6) public release of SynCFM56 with a DOI.
