\chapter{Overcoming the limitations: the F8 program}
\label{ch:overcoming}

\begin{quote}\itshape
This chapter in brief: the limitations declared throughout this document are not footnotes
--- they are the frontier. Each becomes a research line with a discovery hypothesis (the
full ten-line program lives in the project plan). This chapter reports the lines executed
so far, starting with the one that unlocks three others: replacing the linearized generator
with a differentiable neural twin.
\end{quote}

\keyidea{A differentiable neural twin (linearization plus a learned residual) gives full-physics fidelity at microseconds with exact gradients, unlocking the nonlinear fleet --- yet bolting physics features onto a learner fails a second independent way, hardening the warning that ``physics-informed'' is not a free win.}


\section{The program in one view}
\label{sec:f8-program}

Ten declared limitations, ten research lines. The ordering is by synergy, not severity: a
\textbf{differentiable surrogate of the twin} (L1) unlocks nonlinear fleet generation (L2),
the two physics-injection mechanisms H4 left untested --- exact-gradient physics losses and
twin-residual features --- (L6), and fast analysis-by-synthesis diagnosis. Mission-mix
generation (L2) then feeds the report-schedule findings of \cref{ch:tomography} with data
rather than geometry alone. This chapter reports the six lines executed to date: L1 (the twin), L2 (the nonlinear
fleet), L6 (H4 re-opened), L5 (architecture breadth, \cref{sec:f8-l5}), L4 (the recoverable
fraction, \cref{sec:f8-l4}), L9 (validating the PCS metric, \cref{sec:f8-l9}) and L7
(sensor-drift estimation, \cref{sec:f8-l7}). The heavier remaining lines --- real-station
EGT, N-CMAPSS DS02 and map-shape calibration --- queue behind those, each a pre-registrable
study in its own right.
Every line inherits the house discipline: a hypothesis frozen before the run, and a gate that
is allowed to fail.

\section{L1: a differentiable neural twin --- gate passed}
\label{sec:f8-l1}

\paragraph{Why.} The fleet generator is a linearization --- audited to be at the noise
floor for fleet-realistic health states (\cref{sec:audit-nonlin}), but a declared
approximation nonetheless, and one whose error grows exactly where interesting things
happen (end of life, large faults). The full pyCycle model is exact but costs seconds per
snapshot and offers no gradients. The goal: a model with pyCycle's fidelity at
microseconds per evaluation, differentiable end to end.

\paragraph{How.} 2\,400 full pyCycle solves sampled across the design range (all ten
health parameters in $\pm3.5\,\%$, commanded N1 across the cruise band; 8 parallel
workers, zero failed solves, $\sim$14 minutes on the M5). The architecture that passed is
the house pattern at model scale: \textbf{the linearization plus a learned residual} --- a
three-layer MLP corrects what the linear model misses, inheriting its smoothness where
physics is nearly linear. A pure MLP was tried first and could not match the
linearization's precision on N2; the residual form fixed that immediately.

\paragraph{The gate, including its two wrong drafts.} The acceptance gate went through
three disclosed revisions --- each a lesson in evaluation design. Draft one compared the
surrogate's \emph{maximum} error against the linearization's \emph{P95} (a statistic
mismatch). Draft two fixed the statistics but kept the fleet audit's numbers as the
baseline --- the \emph{wrong population}: the audit measured small, fleet-realistic health
states, while the surrogate's test set spans the full $\pm3.5\,\%$ design range, where the
linearization is an order of magnitude worse (N2 nonlinear residual: 0.42\,\% there versus
0.035\,\% in the audit). The final gate evaluates both models \emph{on the same held-out
test set} and demands strict like-for-like improvement.

\begin{table}[htbp]
  \centering\small
  \caption{L1 gate: surrogate versus linearization on the same 400 held-out full-physics
  solves (median and P95 of $|$error$|$, percent of channel value). Auto-generated from
  the surrogate report.}
  \label{tab:surrogate}
  \input{tables/table_surrogate}
\end{table}

\paragraph{Result.} \Cref{tab:surrogate}: the surrogate beats the linearization by
13--19$\times$ on every cockpit channel --- N2 at 0.008\,\% median where the linear model
sits at 0.12\,\%, EGT at 0.03\,\% versus 0.54\,\%. In exchange for 14 minutes of solve
generation, the project now owns a twin that evaluates in microseconds, differentiates
exactly, and is more faithful to the physics than the generator that produced every result
so far --- which is precisely what makes the next lines possible.

\section{L2: the nonlinear fleet (SynCFM56 v2)}
\label{sec:f8-l2}

With both family surrogates gated (\cref{sec:f8-l1}), the fleet generator's physics
emitter becomes swappable: setting \code{fleet.emitter = surrogate} replaces the
linearization of \cref{eq:generator} with the neural twin, at the same
40-second-per-fleet cost (CPU inference inside the existing per-engine workers). Everything
else --- trajectories, sensor model, seeds, splits --- is byte-identical, so v2 differs
from v1.1 in exactly one controlled variable: the fidelity of the true channels.
\textbf{v1.1 is left frozen} (its hashes anchor the F5/F7 pre-registrations); v2 lives
alongside it as \code{data/processed/fleet\_v2/}.

Two re-audits confirm v2. First, fidelity: re-solving full pyCycle at 60 states sampled from
the v2 fleet and comparing both generators against it (\cref{tab:v2-fidelity}), the
surrogate is 3--7$\times$ closer to true physics than the linearization on every cockpit
channel --- EGT median error $0.04\,\%$ versus $0.31\,\%$. Second, the difficulty gate:
the trivial classifier scores $58.0\,\%$ on v2's acute-isolation task (versus $58.1\,\%$
on v1.1), safely below the $60\,\%$ ceiling --- the nonlinear physics did not accidentally
make the benchmark easier, the one risk that mattered. Realism is intact (wash sawtooth
\SI{+2.9}{\degreeCelsius} recovery, end-of-life EGT and fuel-flow signs correct in $100\,\%$ of
engines, measured noise matching the catalog).

\begin{table}[htbp]
  \centering\small
  \caption{L2 fidelity: v1 linear vs v2 surrogate generator error against full pyCycle at
  60 v2-fleet-sampled states ($|$error$|$ as percent of channel value). Auto-generated.}
  \label{tab:v2-fidelity}
  \input{tables/table_v2_fidelity}
\end{table}

\section{L6: H4 re-opened --- and refuted a second way}
\label{sec:f8-l6}

The F5 stacking hybrid failed by concatenating smeared GPA \emph{estimates}
(\cref{sec:res-h4}). L6 tries the mechanism that failure suggested: not the estimate but
the \emph{residual}. Per snapshot, the linear WLS residual
$\mathbf{r} = (\mathbf{I} - \mathbf{H}_{\text{ref}}\mathbf{M})\,\Delta\mathbf{z}$
(with $\mathbf{M}$ the regularized WLS operator, \cref{eq:wls}) is the part of the
measurement the linear GPA \emph{cannot} explain. On v1 this was near zero by construction
(generator and GPA shared the linearization); on v2 the generator is the nonlinear
surrogate, so the residual carries real information --- the twin-residual feature. The RUL
GRU input grows from four channels to seven, and the experiment is frozen exactly as F5's
H4 was (\code{docs/prereg-v3.md}, tag \code{prereg-v3}): pure vs hybrid at 10/25/100\,\%
of training engines, three seeds, single confirmatory pass, thresholds fixed \emph{before}
running.

\begin{table}[htbp]
  \centering\small
  \caption{H4 on v2 (RUL RMSE, cycles, mean of three seeds). The physics feature does not
  help --- it hurts under scarcity. Auto-generated from the frozen verdict.}
  \label{tab:h4v2}
  \input{tables/table_h4v2}
\end{table}

\Cref{tab:h4v2}: the hybrid is \emph{worse} where it was supposed to help most (2\,099 vs
1\,944 cycles at 10\,\% data, 1\,268 vs 1\,233 at 25\,\%) and statistically
indistinguishable at full data; the Wilcoxon test runs the wrong way ($p = 0.99$ against the
hybrid). \textbf{H4 is now refuted a second, independent way.} Two different physics-injection
mechanisms --- estimate-stacking and residual-augmentation --- both fail to improve RUL, on
both the linear and the nonlinear fleet. The transferable finding hardens: for this
prognosis task the raw corrected deviations already contain the remaining-life signal, and
bolting on physics-derived features adds variance, not information. Under the frozen
protocol, mechanism M3 (a physics-constrained loss) is therefore \emph{deferred}: with two
negatives in hand, the burden of proof has shifted, and any future M3 attempt should itself
be pre-registered against this record.

\paragraph{The differentiable twin's real payoff.} L6's negative does not diminish L1: the
surrogate's value is as a fast, exact-gradient \emph{generator} and analysis-by-synthesis
tool (L2; the sensor-drift line is closed in \cref{sec:f8-l7}), not as a feature
factory for a learner that already had the signal.

\section{L5: is the RUL advantage architecture-robust?}
\label{sec:f8-l5}

The headline prognosis result (H3) used a GRU. Is the advantage a property of \emph{learning
from sequences}, or an accident of that one architecture? L5 answers it under the fairness
rule that governs the whole project: three sequence architectures --- the GRU, a temporal
convolutional network (dilated causal convolutions) and a Transformer (self-attention) ---
get \emph{identical} treatment (same inputs, data, RUL cap, epoch budget, three seeds, and
fixed sensible hyperparameters). This is an equal-modest-effort comparison, disclosed as
such (\code{docs/prereg-v8.md}, tag \code{prereg-v8}); it is deliberately \emph{not} a
per-architecture tuning campaign, which would reintroduce the effort asymmetry the project
exists to avoid.

\textbf{H5L.1 confirmed}: all three architectures beat the tuned traditional baseline at
$90\,\%$ life (\cref{tab:arch}), and by nearly the same margin --- RMSE $827$--$888$ cycles
against the traditional $1\,981$, a $\sim$2.3$\times$ improvement whichever network is used.
The convolutional and attention models land within a few percent of the recurrent one; the
best (TCN) beats it slightly, but the finding is precisely that architecture choice barely
matters. The RUL advantage is a property of learning from the deviation sequence, not of the
GRU --- which hardens H3 rather than qualifying it.

\begin{table}[htbp]
  \centering\small
  \caption{L5: RUL RMSE (cycles, mean of three seeds) for three AI architectures under equal
  effort, versus the tuned traditional baseline. Auto-generated.}
  \label{tab:arch}
  \input{tables/table_arch}
\end{table}

\section{L4: how much will a wash recover?}
\label{sec:f8-l4}

A fleet manager washing a compressor wants to know the payoff \emph{before} washing: what
share of an engine's lost EGT margin is fouling (which a wash restores) versus permanent wear
(which it does not). The composable ground truth (\cref{sec:degradation-models}) makes this a
supervised target: replaying each engine's per-mechanism contributions and projecting them
through the takeoff EGT row gives the true recoverable fraction at every cycle. The question
(\code{docs/prereg-v7.md}, tag \code{prereg-v7}): can it be read off the measured deviation
trajectory alone?

\textbf{H4L.1 confirmed, decisively.} A histogram gradient-boosting regressor on
trailing-window shape features (level, slope, curvature and sawtooth roughness of the cockpit
deviations) predicts the recoverable fraction with test $R^2 = 0.86$ and Spearman
$\rho = 0.86$ ($p < 10^{-4}$), against a mean-predictor baseline of zero (\cref{tab:recoverable}).
The mechanism is honest and physical: fouling betrays itself through the wash sawtooth its
recoveries leave in the deviation history, so an engine's own past wash responses reveal how
much the next wash will buy. This is the batch's first clean positive limitation-overcoming
result --- and an actionable one, converting the composability of the synthetic ground truth
into a wash-planning tool an operator could use.

\begin{table}[htbp]
  \centering\small
  \caption{L4: estimating the recoverable (washable) fraction of lost EGT margin from
  deviation-trajectory shape, test split. Auto-generated.}
  \label{tab:recoverable}
  \input{tables/table_recoverable}
\end{table}

\section{L9: can the Physics-Consistency Score be validated?}
\label{sec:f8-l9}

Contribution C5, the Physics-Consistency Score (\cref{sec:pcs}), was a null result on the
weak v0 classifier. L9 asks the fair question under pre-registration
(\code{docs/prereg-v5.md}, tag \code{prereg-v5}): when a classifier reasons more physically,
does PCS rise? Three diagnosis classifiers are scored --- one on the \emph{extended} sensor
set (where the ICM makes faults more separable, \cref{sec:bt-honesty}), one on the cockpit
set (the v0 setting), and a negative control trained on \emph{shuffled} labels
(\cref{tab:pcs}).

\begin{table}[htbp]
  \centering\small
  \caption{L9 PCS validation: isolation accuracy and mean PCS for a competent, a confused,
  and a label-shuffled classifier. Auto-generated.}
  \label{tab:pcs}
  \input{tables/table_pcs}
\end{table}

The verdict is split and, characteristically, deflating in an informative way.
\textbf{H9.2 confirmed}: for the extended classifier PCS is higher on the episodes it gets
right ($0.27$) than on those it gets wrong ($0.24$), so the metric does carry a little
signal about physical correctness. But \textbf{H9.1 is refuted as frozen}: the competent
classifier's PCS ($0.25$) does not significantly exceed the shuffled control's ($0.21$,
Mann--Whitney $p = 0.17$), and the control even edges the honest cockpit classifier. The
reason is the deeper finding, not a bug: \emph{no available classifier is strongly physical},
because acute isolation stays hard even on the extended set (accuracy $0.39$) --- the H2 wall
again. Without a genuinely physical reasoner to validate against, PCS cannot be shown to be
more than weakly informative here. C5 is therefore only partially validated; a clean test
would need a positive control (e.g.\ the extended-sensor WLS estimate, physical by
construction), which is left as pre-registrable future work rather than claimed post hoc.

\section{L7: the lying thermocouple, and the limit of a smarter filter}
\label{sec:f8-l7}

Case C (\cref{ch:cases}) showed a drifting EGT sensor corrupting every downstream method. The
classical remedy is an \emph{augmented-state} Kalman that estimates the sensor bias $b$
alongside the health vector --- the $\mathbf{S}\mathbf{b}$ term of \cref{eq:gpa-linear} made
into a tracked state (\code{docs/prereg-v6.md}, tag \code{prereg-v6}). Two questions: does it
recover the drift, and does it stop the drift from poisoning the diagnosis?

\textbf{It recovers the drift.} The augmented estimate of $b$ tracks the true bias with a
median Spearman $\rho = 0.83$ ($p = 0.004$) across the thirteen drifting-EGT engines. (The
\emph{frozen} confirmatory was restricted to the test split, which contained only two such
engines --- a pre-registration design limitation, disclosed --- so the significant number is
the disclosed exploratory one over all splits; the per-engine recovery is strong either way.)

\textbf{But it barely cleans the diagnosis.} The spurious health error the drift induces on
the hot-section parameters falls only marginally, from $1.69\,\%$ to $1.64\,\%$ ($\sim3\,\%$).
On the cockpit rank-3 measurement set the bias stays confounded with the health directions it
mimics, so the filter can \emph{track} the drift but cannot \emph{un-corrupt} the estimate.
The lesson is the one Case C already implied and the certificate (\cref{ch:breakthrough})
formalizes: against a lying sensor, a smarter algorithm buys awareness, not accuracy ---
the fix is instrumentation (cross-channel redundancy or the extended set), not filtering.

\section{The F8 program so far}
\label{sec:f8-summary}

\Cref{tab:f8-summary} collects the seven executed lines and their verdicts. The shape of the
table is the point: capability-extending lines (L1, L2, L4, L5) succeed, while lines that
probe the confusable-isolation wall (L7, L9) return partial results that reinforce it, and
the one physics-injection line (L6) is a clean negative. The honest ledger is what a research
program looks like when its gates are allowed to fail.

\begin{table}[htbp]
  \centering\small
  \caption{The executed F8 limitation-overcoming lines and their pre-registered verdicts.}
  \label{tab:f8-summary}
  \begin{tabular}{>{\raggedright\arraybackslash}p{0.06\textwidth}>{\raggedright\arraybackslash}p{0.42\textwidth}>{\raggedright\arraybackslash}p{0.44\textwidth}}
    \toprule
    Line & Question & Verdict \\
    \midrule
    L1 & A differentiable twin at pyCycle fidelity? & \textbf{Yes} --- 13--19$\times$ closer than the linearization (\cref{sec:f8-l1}) \\
    L2 & A nonlinear fleet from it? & \textbf{Yes} --- SynCFM56 v2, 3--7$\times$ more faithful, gates re-passed (\cref{sec:f8-l2}) \\
    L5 & Is the RUL win architecture-robust? & \textbf{Yes} --- GRU, TCN, Transformer all beat traditional (\cref{sec:f8-l5}) \\
    L4 & Is the recoverable margin fraction estimable? & \textbf{Yes} --- $R^2 = 0.86$ from trajectory shape (\cref{sec:f8-l4}) \\
    L6 & Does a twin-residual feature help RUL (H4)? & \textbf{No} --- refuted a second, independent way (\cref{sec:f8-l6}) \\
    L7 & Can an augmented Kalman fix a drifting sensor? & \textbf{Partly} --- tracks the drift, cannot un-corrupt the cockpit diagnosis (\cref{sec:f8-l7}) \\
    L9 & Can the PCS metric be validated? & \textbf{Partly} --- tracks correctness weakly; no strongly-physical classifier to test against (\cref{sec:f8-l9}) \\
    \bottomrule
  \end{tabular}
\end{table}
