\chapter{Conclusions and future work}
\label{ch:conclusions}

\begin{quote}\itshape
This chapter in brief: what was built, what the pre-registered evidence says, what an
adopting organization should do differently on Monday, and what remains honestly open.
\end{quote}

\keyidea{What transfers off this simulator is not the absolute numbers but the rankings, mechanisms and geometry --- which method wins where and why, and the observability structure any CFM56-class cockpit set shares. Every operational claim in this document is of that second, transferable kind.}


\section{What this study answers, and how}
\label{sec:questions-answered}

Before the narrative summary, the ledger: every question the study set out to answer, its
answer, and where the evidence lives. The negatives and the partial results are listed with
the same weight as the confirmations --- an honest ledger is one where refutations are as
visible as wins.

\begin{center}\footnotesize
\setlength{\tabcolsep}{4pt}
\textbf{Scientific questions (pre-registered hypotheses).}\\[2pt]
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.30\textwidth}>{\raggedright\arraybackslash}p{0.44\textwidth}>{\raggedright\arraybackslash}p{0.18\textwidth}@{}}
  \toprule
  Question & Answer & Where \\
  \midrule
  Does AI detect faults earlier than trend monitoring? & \textbf{Yes}, once tuned: 12$\times$
    earlier at equal false alarms (recall 0.48 vs 0.13) & \cref{sec:res-h1} (H1) \\
  Does AI isolate confusable faults better? & \textbf{No}: 31\,\%=31\,\%; the wall is
    informational, not algorithmic & \cref{sec:res-h2} (H2) \\
  Does AI predict remaining life better? & \textbf{Yes}: 3--6$\times$ lower error, and robust
    across architectures & \cref{sec:res-h3h5}, \cref{sec:f8-l5} (H3) \\
  Does injecting physics into the learner help? & \textbf{No}, twice: estimate-stacking and
    twin-residual both hurt & \cref{sec:res-h4}, \cref{sec:f8-l6} (H4) \\
  Is the AI's uncertainty better calibrated? & \textbf{Yes}: 6$\times$ tighter interval at
    honest coverage & \cref{sec:res-h3h5} (H5) \\
  Can observability be bought without new hardware? & \textbf{Partly}: report-schedule design
    $+79$\,\%; learned fusion triples classical & \cref{ch:tomography} (H7) \\
  Can we certify what a diagnosis honestly knows? & \textbf{Yes}, validated against truth
    ($\rho=0.70$) & \cref{ch:breakthrough} (H10) \\
  \bottomrule
\end{tabular}
\end{center}

\begin{center}\footnotesize
\setlength{\tabcolsep}{4pt}
\textbf{Operational questions (what an organization can act on).}\\[2pt]
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.30\textwidth}>{\raggedright\arraybackslash}p{0.44\textwidth}>{\raggedright\arraybackslash}p{0.18\textwidth}@{}}
  \toprule
  Question & Answer & Where \\
  \midrule
  Where is AI \emph{not} needed? & Event-shaped faults --- the traditional stack wins at
    near-zero cost & \cref{ch:trad}, \cref{ch:cases} \\
  Which task justifies the AI investment? & Prognosis (the 3--6$\times$, with calibrated
    intervals) & \cref{ch:results} \\
  What can today's sensors know? & Exactly the identifiable subspace the certificate maps &
    \cref{ch:breakthrough} \\
  Which sensor rescues which fault? & Station probes: HPC efficiency $45\times$; isolation
    0.15$\to$0.92 & \cref{sec:bt-honesty}, \cref{sec:f8-lh2} \\
  Can a virtual sensor or clever algorithm substitute a physical one? & \textbf{No} --- it
    adds no information (rank unchanged) & \cref{sec:f8-lh2} \\
  How much will a compressor wash recover? & Estimable from trajectory shape, $R^2=0.86$ &
    \cref{sec:f8-l4} \\
  What is it worth economically? & Median \$3\,M/yr for 100 engines, wide interval, honest
    downside & \cref{ch:economics} \\
  \bottomrule
\end{tabular}
\end{center}

\begin{center}\footnotesize
\setlength{\tabcolsep}{4pt}
\textbf{Method questions (how credible science is done here).}\\[2pt]
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.30\textwidth}>{\raggedright\arraybackslash}p{0.44\textwidth}>{\raggedright\arraybackslash}p{0.18\textwidth}@{}}
  \toprule
  Question & Answer & Where \\
  \midrule
  How to compare AI and traditional without cheating? & Identical data, metrics, splits and
    tuning budget, frozen by pre-registration & \cref{sec:fair-design} \\
  How to trust a synthetic dataset? & Three audits and a difficulty gate that is allowed to
    fail (and did) & \cref{sec:audits} \\
  How to build the benchmark from public data alone? & Certified TCDS/EEDB measurements
    $\rightarrow$ calibrated twin & \cref{ch:model}, \cref{ch:data} \\
  How to prove an uncertainty statement is honest? & Validate it against ground truth ---
    only a synthetic-truth benchmark can & \cref{sec:bt-honesty} \\
  \bottomrule
\end{tabular}
\end{center}

\section{What was built}

From nothing but certified public data, this work produced: a calibrated thermodynamic twin
of the CFM56-7B26 whose fuel-flow predictions sit within 3\,\% of certification test-bed
measurements (C1's foundation); \textbf{SynCFM56}, a 100-engine run-to-failure fleet with
complete per-flight ground truth in airline snapshot format, gated by three audits that
failed and were fixed twice (C1); a traditional EHM pipeline built to industrial shape and a
matched AI suite consuming identical inputs (C2's fairness); conformal uncertainty for
prognosis (C4); the Physics-Consistency Score machinery (C5); the sensor-set observability
analysis (C6's core); an external-benchmark ranking check (C7); and a pre-registered,
Holm-corrected adjudication of five hypotheses under a symmetric 50-trial tuning budget (C2)
--- all reproducible on one desktop machine in minutes and documented to be readable without
prior specialization.

\section{What the evidence says}

Three confirmations, two refutations (\cref{tab:verdicts}), and the split is the message:

\begin{itemize}
  \item \textbf{Learning wins where patterns accumulate}: prognosis (3--6$\times$ lower RUL
    error, calibrated 6$\times$-narrower intervals, ranking replicated on C-MAPSS) and ---
    after honest tuning --- detection timing (12$\times$ earlier at equal false alarms).
  \item \textbf{Physics wins where information is scarce}: the ICM rule matched or beat the
    learner on confusable isolation, and bolting physics-derived features onto a learner
    (H4) made it worse. Physics' real contribution was \emph{diagnosis of the limits}: the
    influence-matrix geometry predicted, before any learning ran, exactly where every method
    would fail.
  \item \textbf{Sensors bound everyone}: the confusable wall (H2) is informational, not
    algorithmic. The extended gas-path set quintuples the worst signature angle; no model
    recovers what probes never measured.
\end{itemize}

\section{Guidance for adoption}

The five-point checklist of \cref{sec:res-industry} stands as the operational summary:
prognosis first, tuned change-detection second, sensors before models for isolation,
distrust of untuned comparisons, and pre-registration as a procurement demand. To it the
case studies add a sixth: \textbf{protect the sensor estate} --- the drifting thermocouple
of Case C corrupts every consumer downstream, learned or classical.

\section{What it is worth}

\Cref{ch:economics} puts a deliberately cautious number on it: for a 100-engine 737NG
operator, the one mechanism the results support --- less-biased prognosis converting
unscheduled removals to scheduled --- plausibly saves low single-digit millions per year
(median \$3\,M, p10 \$0.9\,M), with an honest structural chance of nothing. The larger,
unquantified prize is the diagnostic instrumentation the certificate would justify buying.

\section{Honest limitations}

The world here is synthetic and linearized (audited to the noise floor, but still a model
of a model); the mission mix is two snapshot conditions with scatter; the EGT is a
station-4.5 proxy; chronic labels correlate with age by construction; the AI family is one
compact architecture per task, not the state of the art; and H4's verdict covers one of
three planned hybrid mechanisms. None of these caveats is hidden in fine print --- each is
in the decision register or the relevant chapter.

\section{The capstone: earned-confidence diagnosis}

The project's furthest reach (\cref{ch:breakthrough}) turns its two rarest assets --- a
calibrated differentiable twin and complete ground truth --- into something the field has
wanted and never had: a per-engine certificate of \emph{what gas-path diagnosis can and
cannot know}, proven honest against the true component state ($\rho = 0.70$ between certified
precision and actual error, $p = 0.025$). It converts smearing from a hidden failure into an
explicit, trustworthy, per-direction confidence statement, and the same Fisher-information
object tells an operator which sensor rescues which fault ($45\times$ for HPC efficiency
from station probes) --- unifying the H2 wall and the F7 schedule design under one rigorous
instrument. The honest boundary is stated too: the CRB earns the region's validated anisotropic shape,
and a disclosed split-conformal factor earns its absolute scale (restoring a 90\,\% coverage
guarantee). This is the contribution most likely to matter outside
this document, because it is the one no real-data study can validate.

\section{The F7 extension}

After the main study closed, the operating-point tomography phase (\cref{ch:tomography})
converted the isolation stalemate into two new results under a second pre-registered
freeze: the report schedule as an observability design variable ($+79\,\%$ separability
from one procedural report condition), and learned fusion of physics projections tripling
classical multi-point stacking (0.65 vs 0.17) with calibrated, physics-tracking ambiguity
sets --- plus one honestly refuted threshold hypothesis. The limitations-overcoming program turns every declared limitation into a research line
(\cref{ch:overcoming}). Three are complete: L1 replaced the linearized generator's physics
with a differentiable neural twin 13--19$\times$ more faithful (\cref{fig:surrogate}); L2
regenerated the fleet through it (SynCFM56 v2, 3--7$\times$ closer to full physics,
difficulty gate re-passed); and L6 used v2 to re-open H4 with a twin-residual feature and
\textbf{refuted it a second, independent way} (\cref{fig:h4v2}). Four further lines followed,
each pre-registered: L5 showed the RUL advantage is \emph{architecture-robust} (a TCN and a
Transformer beat the traditional baseline as well as the GRU, \cref{sec:f8-l5}); L4
established that the \emph{recoverable fraction} of lost margin is estimable from trajectory
shape ($R^2 = 0.86$, a wash-planning tool, \cref{sec:f8-l4}); L7 built the augmented-state
Kalman for sensor drift, which tracks the drift but cannot un-corrupt the cockpit diagnosis
(\cref{sec:f8-l7}); and L9 could only partially validate the PCS metric, because no available
classifier is strongly enough physical to test it against (\cref{sec:f8-l9}). The pattern is
the project's in miniature: two clean positives that extend capability, two partial results
that reinforce the observability wall. The three heaviest remaining lines (real-station EGT,
N-CMAPSS DS02, map-shape calibration) are characterized with hypotheses and gates in the
plan.

\section{Future work}

(1) The physics-constrained-loss mechanism (M3), now deferred behind two H4 refutations and
requiring its own pre-registration against that record; (2) the PCS evaluated on the
competent F7 isolation model (C5 validation); (3) N-CMAPSS DS02 as the heavyweight
sim-to-real extension; (4) mission-mix and transient realism in the generator, feeding the
report-schedule design of \cref{ch:tomography} with data rather than geometry; (5) the
extended-sensor fleet, turning the C6 observability analysis into a measured acquisition
trade study; (6) public release of SynCFM56 (v1.1 frozen + v2 nonlinear) with a DOI.
