\chapter{Machine-learning methods in depth}
\label{ch:ml-appendix}

\begin{quote}\itshape
This appendix explains \emph{how} each learning method in the document works, from the ground
up. It assumes only the six-idea primer of \cref{sec:ml-primer}; every method gets a plain
answer to ``what is it?'', an everyday analogy, a small worked example with real numbers
where one helps, and then the mechanism and where it breaks. Read it linearly or dip into the
one method you need; nothing new is claimed here, this is the reference layer the main text
points to.
\end{quote}

\section{How a neural network learns}
\label{sec:ml-nn}

\paragraph{What it is.} A neural network is a mathematical function with many adjustable
numbers (its \emph{parameters}) inside. Training means showing it examples and nudging those
numbers until its outputs match the known answers. Nobody writes the rules; the rules settle
into the numbers.

\paragraph{An analogy.} Think of a mixing desk with thousands of sliders. Each example played
through it comes out slightly wrong; you note which direction each slider should move to make
it less wrong, nudge them all a hair, and repeat over thousands of examples. After enough
passes the desk is tuned. The sliders are the parameters; the ``which direction'' is the
gradient; the ``a hair'' is the learning rate.

\paragraph{The forward pass, concretely.} A layer takes its input numbers, multiplies each by
a weight, adds them up with a bias, and passes the sum through a gentle nonlinearity (here
GELU, essentially ``pass positives, soften negatives''). Stacking layers lets simple pieces
compose into rich input--output relationships. Symbolically one layer is
$\mathbf{a} = \sigma(\mathbf{W}\mathbf{x}+\mathbf{b})$.

\paragraph{A one-neuron worked example.} Take the tiniest network: output
$\hat{y} = w\,x + b$, one weight $w$ and one bias $b$. Suppose the true rule is $y = 2x$, and
we start ignorant at $w=0, b=0$. Show it $x=3$ (so the right answer is $y=6$): it predicts
$\hat{y}=0$, an error of $-6$. The squared-error loss is $\mathcal{L}=(\hat{y}-y)^2=36$. Its
gradient with respect to $w$ is $2(\hat{y}-y)\,x = 2(-6)(3) = -36$: negative, meaning ``$w$ is
too small.'' With learning rate $\eta=0.01$ we update $w \leftarrow 0 - 0.01(-36) = 0.36$.
Repeat over many examples and $w$ climbs toward $2$; the network has \emph{discovered}
the rule $y=2x$ from examples alone. Every network here is this loop, only with millions of
parameters instead of one.

\paragraph{The loss --- the score being minimized.} For predicting a number (remaining life,
a channel value) the loss is the mean squared error, which punishes big misses hard. For
choosing a class (which fault) it is the cross-entropy, which rewards putting probability on
the correct label and punishes confident wrong answers. Rare fault classes are up-weighted so
the network cannot score well by ignoring them.

\paragraph{Backpropagation --- computing all the nudges at once.} A big network has millions
of sliders; computing each one's ``right direction'' separately would be hopeless.
Backpropagation is the chain rule of calculus run backwards through the layers: the error at
the output is passed back, each layer converting it into the nudge for its own weights, in a
single sweep that costs about as much as one forward pass. This one trick is what makes deep
networks trainable.

\paragraph{Gradient descent, mini-batches, epochs.} Each step moves every parameter a small
distance \emph{against} its gradient. The \emph{Adam} optimizer used here gives each parameter
its own adaptive step size (remembering how it has been moving), which trains faster and more
stably than a single fixed rate. Rather than use all data at once, each step uses a random
handful (a \emph{mini-batch}); one full pass through the data is an \emph{epoch}, and training
runs for tens of them.

\paragraph{Why it does not just memorize.} A network with enough parameters can memorize its
training set and fail on anything new --- like a student who learns the practice answers by
heart. The defences here: every reported number comes from engines the network never trained
on (the honest split); training stops after a bounded number of epochs; and the models are
kept small (the RUL network has on the order of ten thousand parameters, not ten million).
Tellingly, bigger networks did not help on this data --- the ceiling is the information in the
measurements, not the size of the model.

\section{The GRU: a network that reads a sequence}
\label{sec:ml-gru}

\paragraph{What it is.} A gated recurrent unit (GRU) is a network that reads a time series one
step at a time, carrying a running summary --- a \emph{hidden state} --- of everything it has
seen. It is the natural tool for an engine's deviation history, which is exactly such a
series.

\paragraph{An analogy.} Picture reading a long report while keeping a single sticky note of
``what matters so far.'' At each paragraph you decide how much of the note to keep and how
much to rewrite with the new information. That running sticky note is the hidden state; the
``how much to keep vs rewrite'' decisions are the GRU's gates.

\paragraph{The problem it solves.} A naive sequence network re-multiplies its summary by the
same matrix at every step. Over hundreds of steps that repeated multiplication shrinks
influences to nothing (or blows them up), so the network forgets the early-life trend by the
time late-life prognosis needs it --- the ``vanishing gradient'' problem. The GRU's
\emph{update gate} can choose to leave the summary \emph{unchanged} for a stretch, letting
information (and its learning signal) survive across long gaps.

\paragraph{The gates, concretely.} At each step the update gate outputs a number between 0 and
1 per feature. If it is $0.1$, the new state is $90\,\%$ old summary and $10\,\%$ fresh
candidate --- memory preserved through a quiet stretch. If it is $0.9$, the state is almost
entirely rewritten --- the network reacting to new evidence. The reset gate similarly controls
how much of the past feeds the fresh candidate. Crucially these gate values are \emph{learned}
and depend on the input, so the network sets its own forgetting rate per feature and per
situation. That is the entire advantage over the classical Holt smoother
(\cref{eq:holt}), whose forgetting rate is one fixed number chosen by hand.

\paragraph{How it is trained.} The network is ``unrolled'' over the window (32--96 steps here)
and backpropagation runs back through every step --- ``backpropagation through time.''
Sequences are cut into fixed windows and down-sampled (every 10th--30th flight) both to bound
the unrolled length and because consecutive flights are nearly identical.

\section{The autoencoder --- and the shortcut that sank it}
\label{sec:ml-ae}

\paragraph{What it is.} An autoencoder is a network trained to \emph{reproduce its own input}
after squeezing it through a narrow middle layer. An \emph{encoder} compresses the input to a
few numbers (the \emph{code}); a \emph{decoder} expands them back. Trained only on healthy
data, it should reconstruct healthy windows well and faulty ones badly --- so a large
reconstruction error flags an anomaly.

\paragraph{An analogy.} It is lossy compression, like a JPEG built only from photographs of
one room: it recreates that room crisply but garbles anything else, and the garble is the
alarm.

\paragraph{Why it failed here (\cref{eq:autoencoder}).} If the middle layer is wide enough,
the cheapest way to minimise reconstruction error is to learn an \emph{identity map} --- copy
the input straight through, like a ``compressor'' that just hands the file back unchanged. It
then reconstructs faulty and healthy windows equally well, and the anomaly score goes flat
(the v0 detector scored zero recall). The lesson is general: a reconstruction detector needs a
bottleneck tight enough to forbid the copy-through shortcut. The Mahalanobis detector below
sidesteps the trap entirely by never reconstructing anything.

\section{Mahalanobis distance: ``how surprising is this reading?''}
\label{sec:ml-maha}

\paragraph{What it is.} A way to measure how far a new measurement lies from normal behaviour,
counted in standard deviations but \emph{accounting for the fact that channels move together}.

\paragraph{An analogy.} Height and weight are correlated: a $190$\,cm person weighing $95$\,kg
is unremarkable, but $190$\,cm at $55$\,kg is very odd even though each number alone is
plausible. Plain distance misses this; the Mahalanobis distance (\cref{eq:mahalanobis})
stretches and rotates the space so that ``normal'' becomes a unit sphere, and then simple
distance in that reshaped space captures true surprise --- the thin, oddly-proportioned case
lands far out.

\paragraph{The one subtlety.} Reshaping the space needs the covariance matrix (how the
channels co-vary). Estimated from limited data it is unreliable in its weakest directions, and
inverting it there amplifies noise into false alarms. \emph{Ledoit--Wolf shrinkage}
(\cref{eq:ledoitwolf}) blends the noisy estimate toward a plain sphere by an amount computed
in closed form --- no knob to tune --- giving a stable, well-behaved distance. An alarm then
requires several consecutive surprising readings, converting a jittery per-flight score into a
detector with a controlled false-alarm rate.

\section{Gradient-boosted trees: a team of rules fixing each other}
\label{sec:ml-gbdt}

\paragraph{What it is.} A decision tree asks a sequence of yes/no questions about the features
(``is the EGT step above $0.4\,\%$? then is the fuel-flow step below\ldots'') and lands each
case in a leaf with a prediction. \emph{Gradient boosting} builds many shallow trees in
sequence, each one trained to fix the errors the previous trees still make
(\cref{eq:gbdt}).

\paragraph{An analogy.} A first rough rule gets many cases right and some wrong; a second rule
studies only the leftover mistakes and patches them; a third patches what those two still miss;
and so on. No single rule is strong, but the accumulating team is. ``Histogram'' boosting just
buckets each feature into a few dozen bins first, so the question-search is fast.

\paragraph{Why trees here.} On small \emph{tabular} problems --- a few hundred cases described
by a handful of engineered numbers, as in fault isolation --- boosted trees are the strongest
off-the-shelf method, which is why isolation used trees while the sequence tasks used the GRU.
Their ceiling here is not the model but the data: on confusable faults the information is
absent (\cref{sec:res-h2}), and no amount of tree depth invents it.

\section{Conformal prediction: honest error bars from any model}
\label{sec:ml-conformal}

\paragraph{What it is.} A wrapper that turns any point prediction into an interval with a
\emph{guaranteed} hit rate --- ``the true value lands inside this band $90\,\%$ of the
time'' --- with no assumption about the model or the noise shape, only that past and future
cases are interchangeable.

\paragraph{A worked example.} Suppose on $19$ held-out engines your RUL predictor's absolute
errors, sorted, are $\{40, 90, 130, \ldots, 1\,900, 2\,600\}$ cycles. For $90\,\%$ coverage
take the error at rank $\lceil 0.9\times(19+1)\rceil = 18$ --- say $1\,900$ cycles --- and call
it $\hat{q}$. Now for a new engine you predict, say, $8\,000$ cycles and report
$8\,000 \pm 1\,900$. The guarantee (\cref{eq:conformal}) holds because a new engine's error is
equally likely to fall at any rank among the old ones, so it exceeds the 18-of-20 mark only
about one time in ten. The ``$+1$'' in the formula is the new point taking its place in the
ranking --- that detail is what makes the guarantee exact for finite data, not just
asymptotically true.

\paragraph{Prediction sets for classification.} The same idea gives a classifier a
\emph{set} of possible labels instead of one guess: accumulate the predicted class
probabilities from most to least likely until a calibrated threshold is reached, and output
that set. When the model is confident the set is a single class; when it is unsure the set
grows --- so ambiguity becomes an explicit, size-measurable, coverage-guaranteed object. This
is exactly what the F7 isolation sets and the F10 certificate (\cref{ch:breakthrough})
exploit.

\section{SHAP: fair credit for a team of features}
\label{sec:ml-shap}

\paragraph{What it is.} A way to split a single prediction into a contribution from each input
feature --- ``the EGT step pushed this call toward \texttt{hpt.eta} by this much, the
fuel-flow step by that much.''

\paragraph{An analogy and its fairness.} Imagine features are teammates and the prediction is
a shared payout. How much did each contribute? SHAP (\cref{eq:shapley}) answers by averaging
each feature's marginal effect over \emph{every order} in which the features could be added to
the team --- the unique split that is provably fair (the parts sum to the whole; two identical
features get equal credit; a feature that never changes the outcome gets nothing). Because
trying every order is expensive, a sampler estimates it. Here the resulting attribution vector
is not the end product but the input to the Physics-Consistency Score
(\cref{sec:pcs}), which asks whether the model's stated reasoning points at the physically
correct component.

\section{Automatic tuning: the tree-structured Parzen estimator}
\label{sec:ml-tpe}

\paragraph{What it is.} An algorithm that chooses the next set of settings to try, learning
from the trials so far instead of scanning a grid blindly.

\paragraph{How it works.} After a few random trials, TPE (\cref{sec:stats-math}) splits them
into ``good'' and ``rest,'' fits a probability distribution over the settings in each, and
proposes the next settings where good outcomes are dense and poor ones are sparse --- steering
toward promising regions while still exploring. On the $14$--$16$-dimensional mixed search
spaces here it finds strong configurations in about $50$ trials, far fewer than a grid would
need --- and, decisive for fairness, both EHM families get the same $50$-trial budget so the
comparison is between two tuned practitioners, not a tuned one and a default one.

\section{The neural surrogate: keep the physics, learn the correction}
\label{sec:ml-surrogate}

\paragraph{What it is.} A network that mimics the slow thermodynamic engine model, predicting
the measured channels from the health state and flight condition in microseconds and with
exact derivatives (\cref{sec:f8-l1}).

\paragraph{The trick that made it work.} Instead of learning the whole map from scratch, it
predicts only the \emph{residual} on top of the F1 linear model:
$\text{surrogate} = \text{linear model} + \text{small network correction}$. The linear model
is already near-perfect where the physics is nearly straight, so the network spends its
capacity only on the curvature the line misses --- like tracing a gentle arc by first drawing
the straight chord and then sketching the small bulge. A from-scratch network could not even
match the linear model on the smoothest channel; the residual form fixed it instantly.
Instructively, the \emph{same} ``keep physics, learn correction'' idea applied to
\emph{features} (not outputs) \emph{failed} for the RUL hybrid (H4): residual learning helps
when the physics is a good first approximation of the target, and hurts when the
physics-derived quantity is just a noisy, redundant extra input to an already-solved problem.

\section{Learned MOPA: let physics invert, let the network fuse}
\label{sec:ml-mopa}

The F7 isolation model (\cref{sec:tomo-learner}) is the clearest lesson in \emph{where}
learning belongs relative to physics. A network fed raw deviation sequences failed at
$12\,\%$ --- asked to re-derive the entire inverse physics from a few hundred examples, it had
far too little data. A network fed the \emph{per-flight physics inversions} (the regularized
least-squares health estimate computed analytically at each flight's condition) and left to do
only the \emph{temporal fusion} of that noisy sequence tripled the classical method. The
division of labour is the moral: hand the known physics the job it does well (inverting the
measurement at one operating point), and hand the network the job it does well (combining many
uncertain estimates over time); neither the all-physics nor the all-network extreme wins.
